Asymmetric relationships and their effects on coexistence

### 2023-08-01 paper/Paper.Rnw
continued trimming and simplifying

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study relationships, ecologists 
very commonly use metrics such as correlation. But standard measures of the strength 
and direction of association between variables do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations are also often used (e.g., various forms of 
regression), but other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples of ATAs in nature have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition have not been investigated. Doing so is our goal.

@@

%Intro to storage effects and why they might be influenced by ATAs
We indicate why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects, starting by conceptually reviewing 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}. 
Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species ($E$) and the strength of competition ($C$) experienced by that species (called $EC$ covariance);
and on the assumption that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive $EC$ covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent $EC$ covariance contributes to 
differences between growth rates of currently rare versus currently common species \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on $EC$ covariation, 
and ATAs can be an aspect of $EC$ covariation. One scenario in which it is reasonable to imagine that
ATAs may be present in $EC$ covariation pertains to water limitation. Suppose two plants compete primarily for 
water, when water is scarce, but their niches become more differentiated when water is abundant.
Then the quality of the environment, $E$, as experienced by whichever plant is scarce, may only correspond closely to the 
competition, $C$, imposed on it by the other, currently abundant plant, when water is scarce and $E$ is low.

@@

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, 
with more details in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, and 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
and 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high. ATA effects quantify how temporal niche partitioning
varies across $EC$ spectrum spanned by the cases 1-2 sketched above.

@@

\subsection{Theoretical setup}

We first develop our new theory for a general two-species model.
Our notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environmental factor $E_i(t)$ and as a decreasing function of 
competitive pressure on species $i$, $C_i(t)$, i.e., $r_i(E_i,C_i)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption is probably unnecessary, it simplifies the
presentation of new ideas.

@@

MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: storage effects. 
So we begin by defining storage effects. 
The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' (GWR) of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species. See \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use, but one definition is $q_{ij}=1$, and to understand the new
ideas presented here it is sufficient to use that definition. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.

@@

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the rare/common species growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of 
tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$. We will refer to this term as 
the \emph{ATA effect on coexistence}. 
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
Note that brackets, $[\cdot]$, and parentheses, $(\cdot)$, are used distinctly in our notation,
throughout the paper - they are not interchangeable.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
See SI section \ref{SI-SIsect:techcomm} for technical comments on the new notation. 
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? Fig. \ref{fig:pedagog2} provides an illustration of the steps we describe here. 
We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal (Fig. \ref{fig:pedagog2}a,b).  
Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\E(e_i)=\E(c_{i \bs i})=0$ and $\var(e_i)=\var(c_{i \bs i})=1$, and we denote 
$\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$ (Fig. \ref{fig:pedagog2}c). We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$ (Fig. \ref{fig:pedagog2}d). 
The value of $\ddot{\rho}_i$ is then determined in one of 
four alternative ways: either by setting $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated.
For those familiar with copula statistics (see, e.g., \citealp{Genest_2007} and \citealp{Ghosh_2020_AER}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The question of the uniqueness of the distribution $(E_i^\nat,C_{i \bs i}^\nat)$ 
is addressed in SI section \ref{SI-SIsect:uniqueness}.
The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created similarly.  
For comparison, Fig. \ref{fig:pedagog2} also shows, for the example explored by that figure, 
the distribution $(E_i^\shp,C_{i \bs i}^\shp)$
defined in the previous section (following \cite{Ellner_2016}).

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, whereas ATAs are 
strictly about covariation. Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}, 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation 
in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT that was goal G1 of the Introduction.

@@

To evaluate the importance of ATAs for a given scenario,
one can quantify
the magnitude of the new ATA effect term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms for the 
scenario.  One typically (but see below) must examine this quantity for each species, $i$,
to obtain complete information about coexistence, since coexistence requires both species populations to 
be able to grow when rare. We also defined scenarios of ``ATA rescue''
and ``ATA exclusion''. \emph{ATA rescue} is considered to have occurred for a given scenario if coexistence
is possible when ATAs are taken into account, i.e., GWR$>0$ for both species, but impossible otherwise, i.e.,
the sum of the non-ATA terms in (\ref{eq:full_decomp}) is negative for at least one species. \emph{ATA exclusion} is 
considered to have occurred if the species do not coexist when ATAs
are taken into account, i.e., GWR$<0$ for at least one species, but would coexist were it not for ATAs, 
i.e., the sum of the non-ATA terms in (\ref{eq:full_decomp})
is positive for both species.

@@

%Setup - two versions, and basics which apply to both
To achieve goal G2 of the Introduction, we applied the framework of Theory to the lottery model, 
comparing the magnitude of the ATA effect on coexistance, $\Delta_i^{[EC]}$, to the magnitudes 
of the other terms of 
the decomposition (\ref{eq:full_decomp}) for various model parameterizations. We used two versions 
of the lottery model. The versions both used the same dynamical equations
(Introduction) and assumed temporally iid fecundities, but the
versions differed in the distributions of $(B_1,B_2)$ used. Both distributions were based on 
transformations of the distributions $(b_1,b_2)$ pictured in Fig. \ref{fig:pedagog1}a-c;
the details of how the left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c), 
and symmetrically (Fig. \ref{fig:pedagog1}b) tail associated alternatives for $(b_1,b_2)$ were generated 
are in SI section \ref{SI-sect:noise}.

>>>>

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study relationships, ecologists 
very commonly use metrics such as correlation. But standard measures of the strength 
and direction of association between variables do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations are also often used (e.g., various forms of 
regression), but other common methods also reveal only a small portion 
of the information content of the relationship \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently 
insights have been gained by studying how features of associations beyond standard measures 
impact ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples of ATAs in nature have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition have not been investigated. Doing so is our goal here.

@@

%Intro to storage effects and why they might be influenced by ATAs
We indicate why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects, starting by conceptually reviewing 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}. 
Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species ($E$) and the strength of competition ($C$) experienced by that species (called $EC$ covariance);
and on the assumption that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive $EC$ covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent $EC$ covariance contributes to 
differences between growth rates of currently rare versus currently common species \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on $EC$ covariation, 
and ATAs can be an aspect of $EC$ covariation. One scenario in which it is reasonable to imagine that
ATAs may be present in $EC$ covariation pertains to water limitation. Suppose two plants compete primarily for 
water, when water is scarce, but their niches become more differentiated when water is abundant.
Then the quality of the environment, $E$, as experienced by whichever plant is currently rare may only correspond closely to the 
competition, $C$, imposed on it by the currently abundant plant when water is scarce and $E$ is low.

@@

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, 
with more details in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, and 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
and 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high. ATA effects quantify how temporal niche partitioning
varies across the $EC$ spectrum spanned by the cases 1-2 sketched above.

@@

\subsection{Setup}

We first develop our theory for a general two-species model.
Our notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environmental factor $E_i(t)$ and as a decreasing function of 
competitive pressure on species $i$, $C_i(t)$, i.e., $r_i(E_i,C_i)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption is probably unnecessary, it simplifies the
presentation of new ideas.

@@

MCT and its computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that only one mechanism relates to ATAs: storage effects. 
So we begin by defining storage effects. 
The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ is the mean ``growth when rare'' (GWR) of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ is the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species. See \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use; but one definition is $q_{ij}=1$, and to understand the new
ideas presented here it is sufficient to use that definition. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.

@@

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which $E$ and $C$
were rendered independent [see (\ref{eq:storage_effects_gen})]. 
Because components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship to the rare/common species growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the $EC$ relationship 
altered. Specifically, ATAs are removed -- see below for how. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the \emph{ATA effect on coexistence}. 
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
Note that brackets, $[\cdot]$, and parentheses, $(\cdot)$, are used distinctly
throughout the paper - they are not interchangeable.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (ATAs).
The symbol should correspondingly be pronounced ``partial sharp''.
See SI section \ref{SI-SIsect:techcomm} for technical comments on the notation. 
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? Fig. \ref{fig:pedagog2} provides an illustration of the steps we describe here. 
We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal (Fig. \ref{fig:pedagog2}a,b).  
Let $\rho_i$ be their covariance. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\E(e_i)=\E(c_{i \bs i})=0$ and $\var(e_i)=\var(c_{i \bs i})=1$, and we denote 
$\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$ (Fig. \ref{fig:pedagog2}c). We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$ (Fig. \ref{fig:pedagog2}d). 
The value of $\ddot{\rho}_i$ is determined in one of 
four alternative ways: either by setting $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated.
For those familiar with copula statistics (see, e.g., \citealp{Genest_2007} and \citealp{Ghosh_2020_AER}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se}. 
The question of uniqueness of the distribution $(E_i^\nat,C_{i \bs i}^\nat)$ 
is addressed in SI section \ref{SI-SIsect:uniqueness}.
The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created similarly.  
For comparison, Fig. \ref{fig:pedagog2} also shows, for the example explored by that figure, 
the distribution $(E_i^\shp,C_{i \bs i}^\shp)$
defined above (following \cite{Ellner_2016}).

@@

The other mechanisms in MCT make no reference to covariation between quantities, whereas ATAs are 
strictly about covariation. Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. The \cite{Ellner_2019} version of MCT comprises 
a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$, as 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation 
in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT that was goal G1 of the Introduction. 
The derivation of this overall decomposition is in SI section 
\ref{SI-SIsect:other_mechanisms}.

@@

To evaluate the importance of ATAs for a given scenario,
we quantified
the magnitude of the new ATA effect term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms for the 
scenario (see Methods). But we also defined scenarios of ``ATA rescue''
and ``ATA exclusion''. \emph{ATA rescue} is considered to have occurred if coexistence
is possible when ATAs are taken into account, i.e., GWR$>0$ for both species, but impossible otherwise, i.e.,
the sum of the non-ATA terms in (\ref{eq:full_decomp}) is negative for at least one species. \emph{ATA exclusion} is 
considered to have occurred if the species do not coexist when ATAs
are taken into account, i.e., GWR$<0$ for at least one species, but would coexist were it not for ATAs, 
i.e., the sum of the non-ATA terms in (\ref{eq:full_decomp})
is positive for both species.

@@

%Setup - two versions, and basics which apply to both
For goal G2 of the Introduction, we applied our theory to the lottery model, 
comparing the magnitude of the ATA effect on coexistance, $\Delta_i^{[EC]}$, to the magnitudes 
of the other terms of 
(\ref{eq:full_decomp}) for various model parameterizations. We used two versions 
of the lottery model. The versions both used the same dynamical equations
(Introduction), but 
differed in the distributions of $(B_1,B_2)$ used. Both distributions were based on 
transformations of the distributions $(b_1,b_2)$ of Fig. \ref{fig:pedagog1}a-c;
the details of how the left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c), 
and symmetrically (Fig. \ref{fig:pedagog1}b) tail associated alternatives for $(b_1,b_2)$ were generated 
are in SI section \ref{SI-sect:noise}.

%the log-normal fecundities version
The \emph{log-normal fecundities lottery model} used $B_i = \exp(\mu_i+\sigma b_i)$,
and therefore had parameters $\sigma$, $\mu_1$ and $\mu_2$ in addition to the death rate parameter $\delta$
(Introduction). The parameter $\sigma$ is the standard deviation of the
log-fecundities and $\mu_i$ is the mean log fecundity for species $i$.
